---
title: "Fossil_Tablet error"
author: "Menghan Fu"
date: "2026-06-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tibble)
library(dplyr)
library(ggplot2)
```

```{r}
baseline_summary <- read.csv("simulation_outputs/baseline_summary.csv")
summary_cv <- read.csv("simulation_outputs/summary_cv.csv")
summary_N1 <- read.csv("simulation_outputs/summary_N1.csv")
summary_ratio <- read.csv("simulation_outputs/summary_ratio.csv")
```


```{r, eval = FALSE}
# Generate target particles
  points <- tibble(
    x = runif(n_target, min = 0, max = region_side),
    y = runif(n_target, min = 0, max = region_side)
  )
  
# Known average marker dose used in the estimator
  known_marker_dose <- N1 * tablet_mean
  
# Generate per-tablet marker doses, then sum them
  tablet_sd <- tablet_mean * tablet_cv
  tablet_doses <- rnorm(N1, mean = tablet_mean, sd = tablet_sd)
  tablet_doses <- pmax(round(tablet_doses), 1)
  
  true_marker_total <- sum(tablet_doses)

# Generate marker particles
  markers <- tibble(
    x = runif(true_marker_total, min = 0, max = region_side),
    y = runif(true_marker_total, min = 0, max = region_side)
  )

```


# Baseline analysis

```{r, eval = FALSE}
set.seed(456)

baseline_result <- simulate_fovs_repeated(
  n_rep = 5000,
  n_target = 10000,
  N1 = 1,
  tablet_mean = 5000,
  tablet_cv = 0,
  region_side = 1,
  fov_size = 0.1,
  n_cal = 20,
  n_ext = 20,
  omega = 2
)

baseline_summary <- baseline_result$summary
```

```{r}
baseline_summary %>%
  select(
    tablet_cv,
    mean_true_conc,
    mean_estimated_conc,
    mean_theoretical_error,
    empirical_cv,
    mean_bias,
    rmse,
    mean_effort,
  ) %>%
  knitr::kable(digits = 2, caption = "Baseline simulation results")
```

# Tablet CV analysis
## Simulation results across tablet CV levels

```{r, eval=FALSE}
tablet_cv_levels <- c(0, 0.01, 0.03, 0.05, 0.07, 0.09, 0.11)

results_cv <- vector("list", length(tablet_cv_levels))

for (i in seq_along(tablet_cv_levels)) {
  results_cv[[i]] <- simulate_fovs_repeated(
    n_rep = 5000,
    n_target = 10000,
    N1 = 1,
    tablet_mean = 5000,
    tablet_cv = tablet_cv_levels[i],
    region_side = 1,
    fov_size = 0.1,
    n_cal = 20,
    n_ext = 20,
    omega = 2
  )$summary
}

summary_cv <- bind_rows(results_cv)
```


```{r}
summary_cv %>%
  select(
    tablet_cv,
    mean_true_conc,
    mean_estimated_conc,
    mean_theoretical_error,
    empirical_cv,
    mean_bias,
    rmse,
    mean_effort,
  ) %>%
  knitr::kable(digits = 2, caption = "Simulation results across tablet CV levels")
```

## Plot: tablet CV and total error

```{r}
ggplot(summary_cv, aes(x = tablet_cv * 100, y = mean_theoretical_error)) +
  geom_line(linewidth = 1, colour = "blue") +
  geom_point(size = 3, colour = "blue") +
  labs(
    title = "Effect of tablet CV on total error",
    x = "Tablet CV (%)",
    y = "Mean total error (%)"
  )
```


## Error decomposition across tablet CV levels

```{r}
summary_cv %>%
  select(
    tablet_cv,
    mean_tablet_term,
    mean_calibration_term,
    mean_marker_term,
    mean_theoretical_error
  ) %>%
  knitr::kable(digits = 2, caption = "Error decomposition across tablet CV levels")
```

```{r}
summary_cv <- summary_cv %>%
  mutate(
    tablet_share = mean_tablet_term^2 / mean_theoretical_error^2,
    calibration_share = mean_calibration_term^2 / mean_theoretical_error^2,
    marker_share = mean_marker_term^2 / mean_theoretical_error^2
  )
```


```{r}
summary_cv_share_long <- summary_cv %>%
  select(
    tablet_cv,
    tablet_share,
    calibration_share,
    marker_share
  ) %>%
  tidyr::pivot_longer(
    cols = c(tablet_share, calibration_share, marker_share),
    names_to = "component",
    values_to = "share"
  )
```


```{r}
ggplot(summary_cv_share_long, aes(x = factor(tablet_cv), y = share, fill = component)) +
  geom_col() +
  labs(
    title = "Relative contributions of error components across tablet CV levels",
    x = "Tablet CV",
    y = "Proportion of total error",
    fill = "Component"
  )
```


# N1 analysis
## Simulation results across numbers of tablets

```{r, eval=FALSE}
N1_levels <- c(1, 2, 4, 6)

total_marker_dose <- 5000
tablet_mean_levels <- total_marker_dose / N1_levels

results_N1 <- vector("list", length(N1_levels))

for (i in seq_along(N1_levels)) {
  results_N1[[i]] <- simulate_fovs_repeated(
    n_rep = 5000,
    n_target = 10000,
    N1 = N1_levels[i],
    tablet_mean = tablet_mean_levels[i],
    tablet_cv = 0.04,
    region_side = 1,
    fov_size = 0.1,
    n_cal = 20,
    n_ext = 20,
    omega = 2
  )$summary
}

summary_N1 <- bind_rows(results_N1)
```

```{r}
summary_N1 %>%
  select(
    N1,
    tablet_cv,
    tablet_mean,
    mean_true_conc,
    mean_estimated_conc,
    mean_theoretical_error,
    empirical_cv,
    mean_bias,
    rmse,
    mean_effort,
  ) %>%
  knitr::kable(digits = 2, caption = "Simulation results across numbers of tablets")
```

## Plot: number of tablets and total error

```{r}
ggplot(summary_N1, aes(x = N1, y = mean_theoretical_error)) +
  geom_line(linewidth = 1, colour = "darkgreen") +
  geom_point(size = 3, colour = "darkgreen") +
  labs(
    title = "Effect of number of tablets on total error",
    x = "Number of tablets (N1)",
    y = "Mean total error (%)"
  )
```

## Error decomposition across numbers of tablets
```{r}
summary_N1 %>%
  select(
    N1,
    mean_tablet_term,
    mean_calibration_term,
    mean_marker_term,
    mean_theoretical_error
  ) %>%
  knitr::kable(digits = 2, caption = "Error decomposition across numbers of tablets")
```

```{r}
summary_N1 <- summary_N1 %>%
  mutate(
    tablet_share = mean_tablet_term^2 / mean_theoretical_error^2,
    calibration_share = mean_calibration_term^2 / mean_theoretical_error^2,
    marker_share = mean_marker_term^2 / mean_theoretical_error^2
  )
```


```{r}
summary_N1_share_long <- summary_N1 %>%
  select(
    N1,
    tablet_share,
    calibration_share,
    marker_share
  ) %>%
  tidyr::pivot_longer(
    cols = c(tablet_share, calibration_share, marker_share),
    names_to = "component",
    values_to = "share"
  )
```


```{r}
ggplot(summary_N1_share_long, aes(x = factor(N1), y = share, fill = component)) +
  geom_col() +
  labs(
    title = "Relative contributions of error components across numbers of tablets",
    x = "Number of tablets (N1)",
    y = "Proportion of total error",
    fill = "Component"
  )
```

# Target-to-marker ratio analysis

$$
u = \frac{n_{\text{target}}}{N_1 \times \mu_{\text{tablet}}}
$$
```{r, eval = FALSE}
set.seed(456)

ratio_levels <- c(1, 2, 4, 6, 8, 10, 15)
n_target <- 10000

results_ratio <- vector("list", length(ratio_levels))

for (i in seq_along(ratio_levels)) {
  current_ratio <- ratio_levels[i]
  current_tablet_mean <- n_target / current_ratio
  
  results_ratio[[i]] <- simulate_fovs_repeated(
    n_rep = 5000,
    n_target = n_target,
    N1 = 1,
    tablet_mean = current_tablet_mean,
    tablet_cv = 0.04,
    region_side = 1,
    fov_size = 0.1,
    n_cal = 20,
    n_ext = 20,
    omega = 2
  )$summary
}

summary_ratio <- bind_rows(results_ratio)
summary_ratio$ratio <- ratio_levels
```


## Simulation results across target-to-marker ratio levels

```{r}
summary_ratio %>%
  select(
    ratio,
    N1,
    tablet_cv,
    tablet_mean,
    mean_true_conc,
    mean_estimated_conc,
    mean_theoretical_error,
    empirical_cv,
    mean_bias,
    rmse,
    mean_effort,
    failure_rate
  ) %>%
  knitr::kable(digits = 2, caption = "Simulation results across target-to-marker ratio levels")
```

## Plot: target-to-marker ratio vs total error

```{r}
ggplot(summary_ratio, aes(x = ratio, y = mean_theoretical_error)) +
  geom_line(linewidth = 1, colour = "purple") +
  geom_point(size = 3, colour = "purple") +
  labs(
    title = "Effect of target-to-marker ratio on total error",
    x = "Target-to-marker ratio",
    y = "Mean total error (%)"
  )
```

## Error decomposition across target-to-marker ratio levels

```{r}
summary_ratio %>%
  select(
    ratio,
    mean_tablet_term,
    mean_calibration_term,
    mean_marker_term,
    mean_theoretical_error
  ) %>%
  knitr::kable(digits = 2, caption = "Error decomposition across target-to-marker ratio levels")
```


```{r}
summary_ratio <- summary_ratio %>%
  mutate(
    tablet_share = mean_tablet_term^2 / mean_theoretical_error^2,
    calibration_share = mean_calibration_term^2 / mean_theoretical_error^2,
    marker_share = mean_marker_term^2 / mean_theoretical_error^2
  )
```


```{r}
summary_ratio_share_long <- summary_ratio %>%
  select(
    ratio,
    tablet_share,
    calibration_share,
    marker_share
  ) %>%
  tidyr::pivot_longer(
    cols = c(tablet_share, calibration_share, marker_share),
    names_to = "component",
    values_to = "share"
  )
```


```{r}
ggplot(summary_ratio_share_long, aes(x = factor(ratio), y = share, fill = component)) +
  geom_col() +
  labs(
    title = "Relative contributions of error components across ratio levels",
    x = "Target-to-marker ratio",
    y = "Proportion of total error",
    fill = "Component"
  )
```

# Presentation slides outline
## Slide 1. Title

- Project title
The effect of tablet error on the precision of estimating the absolute abundance
  of microfossils in the FOVS method
- Name: Menghan Fu
- Student ID: a1941288
- Supervisor: Anthony Mays

## Slide 2. Introduction
What is this project about?

- Specifically focuses on one part of its error structure: tablet error.
- To understand when tablet error matters and when it may be ignored.

Why do we care about fossils (Motivation)?


## Slide 3. Research Questions

- Under what conditions does tablet error become an important part of total FOVS error?
- How does the importance of tablet error change under different experimental designs and sample conditions?
- If tablet error is large, are there design adjustments that can reduce its impact?

## Slide 4. Background

- absolute abundance
- exotic marker method
- FOVS overview

## Slide 5. Background

- FOVS error structure
- tablet error definition
- tablet CV
- research gap

## Slide 6. Methods Overview

How to do the simulation? (Place a simple workflow diagram)

- Set scenario parameters 
- generate fixed-count targets with random locations
- generate random per-tablet marker doses
- sum tablet doses and generate marker locations
- perform calibration and extrapolation counts
- calculate concentration and error components
- repeat simulations and summarise results

## Slide 7. Methods: Scenario Design

Tablet CV scenario

- Fixed: n_target = 10000, N1 = 1, tablet_mean = 5000
- Varied: tablet_cv

N1 scenario

- Fixed: total marker dose = 5000, tablet_cv = 0.04, n_target
- Varied: N1
- tablet_mean = total marker dose/ N1

Ratio scenario

- Fixed: n_target, N1 = 1, tablet_cv = 0.04
- Varied: tablet_mean (This changes the expected target-to-marker ratio)

## Slide 8. Output Metrics

- Mean theoretical error
- empirical_cv
- RMSE
- Mean effort
- Error decomposition
- relative contribution of each error component

## Slide 9. Results: Tablet CV Scenario

Figure:

- total error vs tablet CV
- relative contributions of error components across tablet CV levels

Key message:

- total error increases as tablet CV increases
- tablet_share becomes larger at higher tablet CV
- tablet error can become a dominant source of total error

## Slide 10. Results: N1 Scenario
Figure:

- total error vs N1
- relative contributions across N1

Key message:

- increasing N1 reduces total error
- after revising the scenario to keep total marker dose fixed, increasing N1 mainly reduces tablet-related uncertainty
- marker uncertainty no longer improves automatically

## Slide 11. Results: Ratio Scenario
Figure:

- total error vs ratio
- relative contributions across ratio

Key message:

- total error increases as the target-to-marker ratio increases
- this increase is mainly driven by the marker term
- tablet term stays roughly constant, but its relative contribution declines

## Slide 12. Discussion & Conclusion

- Tablet error is not equally important under all conditions.
- Its importance changes with:
  - tablet dose variation
  - number of tablets
  - target-to-marker ratio
- Increasing the number of tablets can reduce tablet-related uncertainty.
- When tablet CV is high, tablet uncertainty becomes more important.
- When the target-to-marker ratio is high, marker counting uncertainty becomes dominant.



